Tsbrowse: a genome browser for Ancestral Recombination Graphs

### 2025-03-11 manuscript/paper.tex
Pass through the architecture section

\subsection{Architecture} 
\texttt{Tsbrowse} is a client-server application (Fig.~\ref{fig:Figure_1}).
The server is a command-line interface program written in Python, which
is run on the computer where the ARG we wish to visualise is available.
The client is a web-browser, which may optionally be running 
on a different machine. There are two important advantages to this 
approach over a monolithic single-machine architecture. Firstly, 
the ARG being visualised remains in-situ and does not need to be 
downloaded from the server; summaries of the ARG at various 
scales are sent to the client on demand.
Secondly, this provides the flexibility to use a powerful server
when generating summaries of ARG while using a standard desktop 
machine for visualisation.

\texttt{Tsbrowse} is run in a step-wise approach (Fig.~\ref{fig:Figure_1}), 
which makes it convenient to handle large datasets. 
The initial step (\texttt{tsbrowse
preprocess}) prepares the data for visualisation. It can be performed
separately on a large machine, and the output stored for later use. The second
step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates
and renders visualisations. This step follows a client-server architecture. On
the server side (the machine on which \texttt{tsbrowse serve} is run), the
Holoviz suite of libraries \citep{Holoviz} is leveraged to create
visualisations based on the pre-processed data. This choice of architecture is
prompted by the need to represent properties of large-scale ARGs in a
meaningful and performant way. For example, a visual representation of millions
of mutations in SARS-Cov-2 ARGs \citep{zhan2023towards} could quickly
lead to performance bottlenecks and overplotting. Datashader
\citep{Datashader}, a component Holoviz package, overcomes this challenge by
rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
server to render and interact with these visualisations. Heavy data processing
is limited to the server side, keeping the client-side interface lightweight.
For large datasets with complex visualisations, this achieves a smoother user
experience.

\subsection{User Interface}
\label{subsec:User_Interface}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{figures/MainFig2.png}
    \caption{\textbf{A view of ARG edges in \texttt{tsbrowse}.}
    A screenshot of the Edges tab for a 5 Mb sequence simulated with selection coefficient, \textit{s = 0.25}, and beneficial allele situated at \textit{x = 2.5 Mb}. The plot on the left represents each edge as a horizontal line along the sequence (X-axis). The time of parent node is shown on the Y-axis. The persistence of edges across various dimensions is shown through histograms on the right: top, distribution of genomic spans; middle, distribution of time spans; bottom, edge area, which is the product of genomic and time spans.}

    \label{fig:Figure_2}
\end{figure}

\texttt{Tsbrowse}'s interface is built on the \texttt{Holoviz} ecosystem
\citep{Holoviz}, which brings together \texttt{Datashader}, \texttt{Bokeh},
\texttt{Panel}, \texttt{Holoviews} and other libraries to create seamless data
visualisation pipelines even for large datasets. A key challenge with
large-scale data visualisation is to maintain a responsive user interface.
\texttt{Datashader} aggregates data and renders only the necessary data points
at a given zoom level, thereby minimising the computational load on the
browser. A \texttt{Bokeh} plotting back-end provides useful tools and widgets.
The \textit{pan}, \textit{select} and \textit{zoom} tools enable interactivity
with individual data points, allowing users to gain deeper insights from the
data. The \texttt{serve} step accepts an optional input file containing
sequence annotations. This allows users to overlay contextual information about
genes or other sequence features.

The user interface is presented as a \texttt{Panel} dashboard, with pages to
describe mutations, nodes, edges and trees in the input ARG. Figure 2
demonstrates the user interface using the Edges page as an example. The plot on
the left is an overview of all 33,929 edges (parent-child relationships between
node-pairs; see \hyperref[subsec:Data_Model]{Data Model}) in a simulated ARG
\hyperref[subsec:sweep_simulation]{described in Supplementary Methods}; each
edge is depicted as a horizontal line connecting the genomic coordinates of the
parent and child nodes on the x-axis. The y-axis denotes the age of either
node. A useful feature of the app is the ability to interact with the
visualisations. The tools on the right of the plot allow the user to examine
edges within a sequence/time slice, and query individual edges. The histograms
to the right provide summaries of edge spans over sequence and time.

>>>>

\subsection{Architecture} 
\texttt{Tsbrowse} is a modular client-server application 
written in Python, optimised for handling ARGs with millions
of samples (Fig.~\ref{fig:Figure_1}). 
To ensure that such large ARGs can be visualised quickly
we perform a preprocessing step (\texttt{tsbrowse preprocess})
which augments the \texttt{tskit} 
data model with some additional derived information (e.g., the number 
of inheritors of each mutation).
To visualise an ARG we then run \texttt{tsbrowse serve}, which 
by default will open a browser window on the local machine, but 
also supports running the server on a remote machine across the 
network.
This client-server architecture has some important advantages over
a monolithic single-machine approach. Most importantly,
the ARG being visualised remains in-situ and does not need to be 
downloaded from the server.

The express goal of \texttt{tsbrowse} is to provide interpretable
and interactive visual summaries of ARGs containing millions of nodes, edges 
and mutations. Mutations, for example, have a clear interpretation
when plotted on genome coordinate vs time axes, but at this scale 
the data density is far too high to simply plot each mutation as a point.
We overcome this problem by using Datashader and the wider 
Holoviz ecosystem~\citep{Holoviz}, which efficiently rasterises large datasets
at the requested resolution on the server, sends the image to the web-browser
for display, and dynamically updates as the user interatively navigates the ARG.

% and renders visualisations. This step follows a client-server architecture. On
% the server side (the machine on which \texttt{tsbrowse serve} is run), the
% Holoviz suite of libraries \citep{Holoviz} is leveraged to create
% visualisations based on the pre-processed data. This choice of architecture is
% prompted by the need to represent properties of large-scale ARGs in a
% meaningful and performant way. For example, a visual representation of millions
% of mutations in SARS-Cov-2 ARGs \citep{zhan2023towards} could quickly
% lead to performance bottlenecks and overplotting. Datashader
% \citep{Datashader}, a component Holoviz package, overcomes this challenge by
% rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
% for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
% tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the

% \texttt{Tsbrowse} is run in a step-wise approach (Fig.~\ref{fig:Figure_1}), 
% which makes it convenient to handle large datasets. 
% The initial step (\texttt{tsbrowse
% preprocess}) prepares the data for visualisation. It can be performed
% separately on a large machine, and the output stored for later use. The second
% step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates
% and renders visualisations. This step follows a client-server architecture. On
% the server side (the machine on which \texttt{tsbrowse serve} is run), the
% Holoviz suite of libraries \citep{Holoviz} is leveraged to create
% visualisations based on the pre-processed data. This choice of architecture is
% prompted by the need to represent properties of large-scale ARGs in a
% meaningful and performant way. For example, a visual representation of millions
% of mutations in SARS-Cov-2 ARGs \citep{zhan2023towards} could quickly
% lead to performance bottlenecks and overplotting. Datashader
% \citep{Datashader}, a component Holoviz package, overcomes this challenge by
% rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
% for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
% tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
% server to render and interact with these visualisations. Heavy data processing
% is limited to the server side, keeping the client-side interface lightweight.
% For large datasets with complex visualisations, this achieves a smoother user
% experience.

\subsection{User Interface}
\label{subsec:User_Interface}
\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{figures/MainFig2.png}
    \caption{\textbf{A view of ARG edges in \texttt{tsbrowse}.}
    A screenshot of the Edges tab for a 5 Mb sequence simulated with selection coefficient, \textit{s = 0.25}, and beneficial allele situated at \textit{x = 2.5 Mb}. The plot on the left represents each edge as a horizontal line along the sequence (X-axis). The time of parent node is shown on the Y-axis. The persistence of edges across various dimensions is shown through histograms on the right: top, distribution of genomic spans; middle, distribution of time spans; bottom, edge area, which is the product of genomic and time spans.}

    \label{fig:Figure_2}
\end{figure}
% \texttt{Tsbrowse}'s interface is built on the \texttt{Holoviz} ecosystem
% \citep{Holoviz}, which brings together \texttt{Datashader}, \texttt{Bokeh},
% \texttt{Panel}, \texttt{Holoviews} and other libraries to create seamless data
% visualisation pipelines even for large datasets. A key challenge with
% large-scale data visualisation is to maintain a responsive user interface.
% \texttt{Datashader} aggregates data and renders only the necessary data points
% at a given zoom level, thereby minimising the computational load on the
% browser. A \texttt{Bokeh} plotting back-end provides useful tools and widgets.
% The \textit{pan}, \textit{select} and \textit{zoom} tools enable interactivity
% with individual data points, allowing users to gain deeper insights from the
% data. The \texttt{serve} step accepts an optional input file containing
% sequence annotations. This allows users to overlay contextual information about
% genes or other sequence features. 
The user interface is presented as a dashboard, with pages to
describe mutations, nodes, edges and trees in the input ARG. Figure 2
demonstrates the user interface using the Edges page as an example. The plot on
the left is an overview of all 33,929 edges (parent-child relationships between
node-pairs; see \hyperref[subsec:Data_Model]{Data Model}) in a simulated ARG
\hyperref[subsec:sweep_simulation]{described in Supplementary Methods}; each
edge is depicted as a horizontal line connecting the genomic coordinates of the
parent and child nodes on the x-axis. The y-axis denotes the age of either
node. A useful feature of the app is the ability to interact with the
visualisations. The tools on the right of the plot allow the user to examine
edges within a sequence/time slice, and query individual edges. The histograms
to the right provide summaries of edge spans over sequence and time.
